The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: some prose changes, structurally ready

The type $\Xcomp{m}{n}[D=E_0,\ldots,E_{n-1}]$ of $E_n$ represents a family whose type can be presented into the form $\fullframe(E_0,\ldots,E_{n-1}) \rightarrow \U$ where $\fullframe$ collects all the components on which $E_n$ is expected to depend. It is recursively defined in table~\ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. A $\fullframe$ describes a boundary of a standard form (simplex, cube), which we decompose into $\layer$, and a $\painting$ corresponds to a filled frame. Notice that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in table~\ref{tab:faces}, and this restriction operator is in turn defined using auxiliary definitions $\restrl$ and $\restrp$.

@@

In \citep{hr25}, the well-foundedness was justified by induction on $n$ and subinductions, one on $p$ to mutually build $\framep[n][p]$, $\restrf[n][p]$ and $\cohf[n][p]$, and three others to build each of $\painting[n][p]$, $\restrp[n][p]$ and $\cohp[n][p]$ by induction on $n-p$.

In the current paper and associated formalisation, we take $\framep[n-2][p]$ $\painting[n-2][p]$, $\restrf[n-1][p]$ and $\restrp[n-1][p]$ as references and rename them respectively as $\framep[n][p]$, $\painting[n][p]$, $\restrf[n][p]$ and $\restrp[n][p]$. We replace the equational properties of $\framep[n][p]$, $\framep[n-1][p]$, $\restrf[n][p]$, $\painting[n][p]$, $\painting[n-1][p]$ and $\restrp[n][p]$ with definitions, renaming them respectively to $\framep[n+2][p]$, $\framep[n+1][p]$, $\restrf[n+1][p]$, $\painting[n+2][p]$, $\painting[n+1][p]$ and $\restrp[n+1][p]$. This means that we build instead by induction a block of the form:

\begin{equation*}
  \begin{array}{lclr}
    \framep[n][p]   & : & \U                                                                                                  & \mbox{for all $p\leq n$} \\
    \painting[n][p] & : & \framep[n][p] \rightarrow \U                                                                        & \mbox{for all $p\leq n$} \\
    \restrf[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p] \rightarrow \framep[n][p]                         & \mbox{for all $p\leq n$} \\
    \restrp[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \Pi d:\framep[n+1][p].\,                                                                     \\
                    &   & \painting[n+1][p](d) \rightarrow \painting[n][p](\restrf[n][p](d))                                  & \mbox{for all $p\leq n$} \\
    \cohf[n][p]     & : & \Pi qr\epsilon\omega.\, \restrf[n][p][q][\epsilon] \circ \restrf[n+1][p][\omega][r] =                                          \\
                    &   & \restrf[n][p][\omega][r] \circ \restrf[n+1][p][q+1][\epsilon]                                       & \mbox{for all $p\leq n$} \\
    \cohp[n][p]     & : & \Pi qr\epsilon\omega.\,\Pi d:\framep[n+2][p].\,                                                                                \\
                    &   & \restrp[n][p][q][\epsilon](\restrf[n+1][p][r][\omega](d)) \circ \restrp[n+1][p][r][\omega](d) =                                \\
                    &   & \restrp[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon](d)) \circ \restrp[n+1][p][q+1][\epsilon](d) & \mbox{for all $p\leq n$} \\
  \end{array}
\end{equation*}

Turning $\framep[n+2][p]$, $\framep[n+1][p]$ and $\restrf[n+1][p]$ into definitions involves however circularities. To start with, the definition of $\framep[n+1][p]$, which is by induction on $p$, depends on $\restrf[n][p-1]$ whose type depends in turn on $\framep[n+1][p-1]$. This means that $\framep[n+1][p]$ and the type of $\restrf[n][p]$ have to be mutually defined.

The definition of $\framep[n+1][p]$ also depends, via $\layer[n][p-1]$, on $\painting[n][p-1]$. Both via $\painting[n][p-1]$ and $\restrf[n][p-1]$, it depends also on $\framep[n][p-1]$ by typing dependency. This requires to assume $\framep[n][p-1]$ and $\painting[n][p-1]$ known in advance.
Moreover, the definition of $\framep[n+1][p]$ depends on $\framep[n+1][p-1]$, so all of $\framep[n][k]$, $\painting[n][k]$ and $\restrf[n][k]$ for $k< p$ need to be known, which we respectively abbreviate as $\framep[n][\lbrack 0,p-1 \rbrack]$, $\painting[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][p-1]$.
Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following dependencies for $p\leq n+1$ fixed:
\begin{center}
  \begin{tikzcd}
    \framep[n+1][\lbrack 0,p \rbrack] \arrow[d] \arrow[ddr]\\
    \painting[n][\lbrack 0,p-1 \rbrack] \arrow[d, dotted] \\
    \framep[n][\lbrack 0,p-1 \rbrack] &
    \restrf[n][\lbrack 0,p-1 \rbrack] \arrow[l, dotted] \arrow[uul, dotted, "\mbox{$[0,p-1]$}"', near start, shift right=2mm, hook']\\
  \end{tikzcd}
\end{center}

@@

Note that we write $\framep$, $\painting$, $\restrf$ for the
projections of a $\deps$. Also, by abuse of notation, we write $\unitpoint$ for
the only inhabitant of $\depstype[n]$.

@@

TODO: say the convention on projection in the Introduction

@@

We need also the notion of full dependencies at the level of coherence conditions, that combines a full $\depscohs[n][\lbrack 0,n \rbrack]$ with a set $E:\framep[n+2][n+2](\overline{\deps}(\depscohs[n][\lbrack 0,n \rbrack])) \rightarrow \U$:
\begin{equation*}
  \fulldepscohstype[n] \defeq
  \left(\begin{array}{lcl}
      \depscohs[n][\lbrack 0,n-1 \rbrack] & : & \depscohstype[n][\lbrack 0,n \rbrack]                                                \\
      E                                   & : & \framep[n+2][n+2](\overline{\deps}(\depscohs[n][\lbrack 0,n \rbrack]))\rightarrow \U \\
    \end{array}\right)
\end{equation*}
TODO: say the convention on projection in the Introduction

>>>>

The type $\Xcomp{m}{n}[D=E_0,\ldots,E_{n-1}]$ of $E_n$ represents a family whose type can be presented into the form $\fullframe(E_0,\ldots,E_{n-1}) \rightarrow \U$ where $\fullframe$ collects all the dependencies of $E_n$. It is recursively defined in table~\ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. A $\fullframe$ describes a boundary of a standard form (simplex, cube), which we decompose into $\layer$ themselves made of $\nu$ $\painting$ ($\nu=1$, $\nu=2$, ...), where a $\painting$ corresponds to a filled frame. Notice that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in table~\ref{tab:faces}, and this restriction operator is in turn defined using auxiliary definitions $\restrl$ and $\restrp$ (see \cite{hr25} for details).

@@

To build $\framep[n+1]$ from $\frame[n]$, it happens to be enough to assume the following built at level $n$:

\begin{equation*}
  \begin{array}{lclr}
    \framep[n][p]   & : & \U                                                                                                  & \mbox{for all $p\leq n$} \\
    \painting[n][p] & : & \framep[n][p] \rightarrow \U                                                                        & \mbox{for all $p\leq n$} \\
    \restrf[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \framep[n+1][p] \rightarrow \framep[n][p]                         & \mbox{for all $p\leq n$} \\
    \restrp[n][p]   & : & \Pi q \leq n-p.\, \Pi \epsilon.\, \Pi d:\framep[n+1][p].\,                                                                     \\
                    &   & \painting[n+1][p](d) \rightarrow \painting[n][p](\restrf[n][p](d))                                  & \mbox{for all $p\leq n$} \\
    \cohf[n][p]     & : & \Pi qr\epsilon\omega.\, \restrf[n][p][q][\epsilon] \circ \restrf[n+1][p][\omega][r] =                                          \\
                    &   & \restrf[n][p][\omega][r] \circ \restrf[n+1][p][q+1][\epsilon]                                       & \mbox{for all $p\leq n$} \\
    \cohp[n][p]     & : & \Pi qr\epsilon\omega.\,\Pi d:\framep[n+2][p].\,                                                                                \\
                    &   & \restrp[n][p][q][\epsilon](\restrf[n+1][p][r][\omega](d)) \circ \restrp[n+1][p][r][\omega](d) =                                \\
                    &   & \restrp[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon](d)) \circ \restrp[n+1][p][q+1][\epsilon](d) & \mbox{for all $p\leq n$} \\
  \end{array}
\end{equation*}
and to build it at level $n+1$. This is however not so easy because some components of this structure, e.g. $\restrf$ and further fields, depend on $\framep[n+1]$, that is already depend on the ability of building the $\framep$ component of the structure at level $n+1$. We are thus facing an interdependency between specifying the structure assumed at level $n$ and defining from it the structure at level $n+1$. So, we have to proceed in order: specifying $\framep[n]$, from which $\painting[n]$ can be specified, from which $\framep[n+1]$ can be built and thus $\restrf[n+1]$ be specified. Now, the definition of $\framep[n+1][p]$, which is by induction on $p$, itself depends on $\restrf[n][p-1]$ whose type depends in turn on $\framep[n+1][p-1]$. This means that $\framep[n+1][p]$ and the type of $\restrf[n][p]$ have to be mutually defined.

The definition of $\framep[n+1][p]$ also depends, via $\layer[n][p-1]$, on $\painting[n][p-1]$. Both via $\painting[n][p-1]$ and $\restrf[n][p-1]$, it depends also on $\framep[n][p-1]$ by typing dependency. This requires to assume $\framep[n][p-1]$ and $\painting[n][p-1]$ known in advance. Moreover, the definition of $\framep[n+1][p]$ depends on $\framep[n+1][p-1]$, so all of $\framep[n][k]$, $\painting[n][k]$ and $\restrf[n][k]$ for $k< p$ need to be known, which we respectively abbreviate as $\framep[n][\lbrack 0,p-1 \rbrack]$, $\painting[n][\lbrack 0,p-1 \rbrack]$ and $\restrf[n][\lbrack 0,p-1 \rbrack]$. Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following dependencies for $p\leq n+1$ fixed:

\begin{center}
  \begin{tikzcd}
    \framep[n+1][\lbrack 0,p \rbrack] \arrow[d] \arrow[ddr]\\
    \painting[n][\lbrack 0,p-1 \rbrack] \arrow[d, dotted] \\
    \framep[n][\lbrack 0,p-1 \rbrack] &
    \restrf[n][\lbrack 0,p-1 \rbrack] \arrow[l, dotted] \arrow[uul, dotted, "\mbox{$[0,p-1]$}"', near start, shift right=2mm, hook']\\
  \end{tikzcd}
\end{center}

@@

Note that $\deps$ behaves here as an abbreviation for a sequence of assumptions. In particular, once a $\deps:\depstype$ is assumed, we can freely refer to $\framep$, $\painting$, $\restrf$. Also, by abuse of notation, we write $\unitpoint$ for the only inhabitant of $\depstype[n][0]$.

@@

@@

We need also the notion of full dependencies at the level of coherence conditions, that combines a full $\depscohs[n][\lbrack 0,n \rbrack]$ with a set $E:\framep[n+2][n+2](\overline{\deps}(\depscohs[n][\lbrack 0,n \rbrack])) \rightarrow \U$:
\begin{equation*}
  \fulldepscohstype[n] \defeq
  \left(\begin{array}{lcl}
      \depscohs[n][\lbrack 0,n-1 \rbrack] & : & \depscohstype[n][\lbrack 0,n \rbrack]                                                \\
      E                                   & : & \framep[n+2][n+2](\overline{\deps}(\depscohs[n][\lbrack 0,n \rbrack]))\rightarrow \U \\
    \end{array}\right)
\end{equation*}
